Power to the Stimuli: Not the Effect

### 2023-03-29 manuscript_draft.tex
so much all the updates

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Power to the Stimuli: Not the Effect},
  pdfauthor={Erin M. Buchanan1 \& Other Folks as Per Order on Doc2},
  pdflang={en-EN},
  pdfkeywords={power, sampling, accuracy in parameter estimation},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

@@

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\abstract{%
We will add an abstract that explains that we are going to 1) talk about this cool procedure, 2) show some simulations that the procedure works, and 3) give two examples of the procedure in action. You should try it out!
}

@@

This part we will put in a real introduction. Things we need to discuss:

\begin{itemize}
\tightlist
\item
  Repeated measures data with items
\item
  Accuracy in Parameter Estimation
\item
  Adaptive sampling
\item
  MLM designs account for item heterogeneity but a lot of people anova
\end{itemize}

@@

Using ideas from accuracy in parameter estimation, we suggest the following procedure to determine a sample size for each item:

@@

\emph{Samples}. Each population was then sampled as if a researcher was conducting a pilot study. The sample sizes started at 20 participants per item increasing in units of 5 up to 100 participants.

@@

In this section, we simulate what a researcher might do if they follow our suggested application of AIPE to sample size planning based on well measured items. Assuming each pilot sample represents a dataset a researcher has collected, we will simulate samples of 20 to 500 to determine what the new sample size suggestion would be. We assume that samples over 500 may be considered too large for many researchers who do not work in teams or have participant funds. The standard error of each item was calculated for each suggested sample size by pilot sample size by population type.

@@

@@

\(N_{Pilot}\) indicates the sample size of the pilot data minus the minimum simulated sample size to ensure that the smallest sample sizes do not decay (i.e., the formula zeroes out). This value is raised to the power of \(log_2\) of the sample size of the pilot data, which decreases the impact of the decay to smaller increments for increasing sample sizes. This value is then multiplied by the projected sample size. As show in Figure \ref{fig:corrected-figure}, this correction factor produces the desired quality of maintaining that small pilot studies should \emph{increase} sample size, and that sample size suggestions level off as pilot study data sample size increases.

@@

To explore if we could recover the new projected sample size from data a researcher would have, we used linear models to create a formula for researcher correction. First, the corrected projected sample size was predicted by the original projected sample size. Next, the standard deviation of the item standard deviations was added to the equation to recreate heterogeneity estimates. The scale of the data is embedded into the standard deviation of the items (\emph{r} = 0.81), and therefore, this variable was not included separately. Last, we included the pilot sample size.

The first model using pilot sample size to predict new sample size was significant, \emph{F}(1, 5666) = 48,042.64, \emph{p} \textless{} .001, \(R^2 = .89\), 90\% CI \([0.89, 0.90]\), capturing nearly 90\% of the variance, \(b = 0.62\), 95\% CI \([0.62, 0.63]\). The second model with item standard deviation was better then the first model \emph{F}(1, 5665) = 55.21, \emph{p} \textless{} .001, \(R^2 = .89\), 90\% CI \([0.89, 0.90]\). The item standard deviation predictor was significant, \(b = 0.02\), 95\% CI \([0.01, 0.03]\), \(t(5665) = 4.54\), \(p < .001\). The addition of the original pilot sample size was also significant, \emph{F}(1, 5664) = 9,529.83, \emph{p} \textless{} .001, \(R^2 = .96\), 90\% CI \([0.96, 0.96]\).

@@

\begin{tabular}{lllll}
\toprule
Term & Estimate & $SD$ & $t$ & $p$\\
\midrule
Intercept & 39.269 & 0.437 & 89.843 & < .001\\
Projected Sample Size & 0.700 & 0.002 & 366.669 & < .001\\
Item SD & 0.003 & 0.003 & 0.952 & .341\\
Pilot Sample Size & -0.695 & 0.007 & -97.621 & < .001\\
\bottomrule
\end{tabular}

@@

Last, we examine the question of an appropriate SE decile. All graphs for power, heterogeneity, scale, and correction are presented online. First, the 0\%, 10\%, and 20\% deciles are likely too restrictive, providing very large estimates that do not always find a reasonable sample size in proportion to the pilot sample size, scale, and heterogeneity. If we examine the \(R^2\) values for each decile of our regression equation separately, we find that the 50\% (0.96) represents the best match to our corrected sample size suggestions. The 50\% decile, in the corrected format, appears to meet all goals: 1) increases with heterogeneity and scale of data, and 2) higher suggested values for small original samples and a leveling effect at larger pilot data. Figure \ref{fig:decile-figure} illustrates the corrected scores for simulations at the 50\% decile recommended cutoff for item standard errors.

@@

In this section, we provide two examples of the suggested procedure. The first example includes concreteness ratings from Brysbaert et al.~{[}CITE, 2014{]}. Instructions given to participants denoted the difference between concrete (i.e., ``refers to something that exists in reality'') and abstract (i.e., ``something you cannot experience directly through your senses or actions'') terms. Participants were then asked to rate concreteness of terms using a 1 (abstract) to 5 (concrete) scale. This data represents a small scale dataset that could be used as pilot data for a study using concrete word ratings. The data is avaliable at OSF LINK {[}CITE{]}. The second dataset includes a large scale dataset with response latencies, the English Lexicon Project {[}CITE, Balota et al.{]}. The English Lexicon Project consists of lexical decision response latencies for English words. In a lexical decision task, participants simply select ``word'' for real words (e.g., \emph{dog}) and ``nonword'' for pseudowords (e.g., \emph{wug}). The trial level data is avaliable here {[}\url{https://elexicon.wustl.edu/}, CITE{]}. Critically, in each of these examples, the individual trial level data for each item is available to simulate and calculate standard errors on. Data that has been summarized could potentially be used, as long as the original standard deviations for each item were present. From the mean and standard deviation for each item, a simulated pilot dataset could be generated for estimating new sample sizes. All code to estimate sample sizes is provided on our OSF page.

@@

The concreteness ratings data includes 63039 concepts that were rated for their concreteness. In our fictional study for this example, we selected 100 random words to show participants. In the original study, not every particpant rated every word, which created uneven sample sizes for each word. In our random sample of 100 words, the average pilot sample size was 27.96 (\emph{SD} = 1.47), and we will use 28 as our pilot sample size for this example. All ``do not know'' ratings were set as missing data. The 50\% decile for items standard error was 0.25 for our cutoff criterion.

The pilot data was then simulated, with replacement, with samples from 20 to 300 increasing in units of 5. On each sample, the percent of items below the cutoff score were calculated. After applying our correction equation, we find that a sample size of 44 would allow for at least 80\% of items to meet the cutoff criterion. The sample sizes for 85\% (48), 90\% (48), and 95\% (51) are also options for sample size suggestions. Finally, we calculated the potential amount of data retention given that participants could indicate they did not know a word (\(M_{correct}\) = 0.82, \emph{SD} = 0.24). In order to account for this facet, the potential sample sizes were multiplied by 1/0.82, which results in a suggested sample of 54, 59, and 63. Therefore, we could designate our minimum sample per item as 54, stopping rule of 0.25, and maximum sample size of 63.

@@

The ELP response latency data includes 80962 word-forms, 40481 that are listed as non-words, and 40481 real words. For our example study, we will randomly select 500 real words and 500 non-words to show participants. The average pilot sample size for this random sample was 32.68 (\emph{SD} = 0.64), and \emph{n} = 33 will be our pilot size for this example. Again, participants are expected to make mistakes, and we calculated percent correct as 0.86, which was roughly even in the two stimulus categories: \(M_{word}\) = 0.84 and \(M_{non-word}\) = 0.87. The 50\% decile for items standard error was 61.32 for our cutoff criterion. We additionally checked to ensure that the two stimulus types did not have very different cutoff criterions: 50\% decile \(SE_{words}\) = 56.94, 50\% decile \(SE_{nonwords}\) = 65.61. In this scenario, we could chose to go with the lower SE to be more conservative (i.e., higher projected sample size). Given the values were close for large scale data, we used the 50\% decile of all stimuli taken together.

The pilot response latency data was then simulated in the same way as described above. After calculating the percent below our cutoff score, we applied the correction to the projected sample sizes. A sample size of 31 would equate to 80\% of the items reaching our cutoff, along with 85\% (34), 90\% (34), and 95\% (38). Again, we adjusted for data loss given that participants are expected to incorrectly answer items, resulting in a suggested sample of 36, 40, and 44. One other possible consideration for this study is potential fatigue in showing participants 1000 target items. Therefore, we could designate in our research design that each participant will only receive 500 of the target items. We would need to double our sample sizes to account for splitting of the items across multiple sets of participants. Our minimum sample size for the entire study could be 72, stopping rule of 61.32, and maximum sample size of 88. This study would benefit from an adaptive design, where smaller sets items are randomly sampled for participants until they reach the minimum sample size or the cutoff criteria. At this point, items are probablistically sampled (e.g., higher selection probability for items that have not reached a minimum or stopping rule) until all items have reached criteria.

@@

While the examples in this manuscript are traditionally cognitive linguistics focused, any research using repeated items can benefit from newer sampling techniques. Therefore, we provide XX example vignettes on our OSF page/GitHub site for this manuscript across a range of examples of data types provided by the authors of this manuscript. Examples include \ldots{} ADD HERE AFTER DONE WITH VIGNETTES.

@@

To be added once people say this idea doesn't suck.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\makeatletter
\usepackage{etoolbox}
\patchcmd{\maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

@@

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Power to the Stimuli: Not the Effect},
  pdfauthor={Erin M. Buchanan1 \& Other Folks as Per Order on Doc2},
  pdflang={en-EN},
  pdfkeywords={power, sampling, accuracy in parameter estimation},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\abstract{%
Sample size planning for research studies often focuses on obtaining a significant result given a specified level of power, alpha, and proposed effect size. This planning generally requires prior knowledge of study design and a statistical analysis to calculate the proposed sample size. However, there may not be just one specific testable analysis from which to derive power (Silberzahn et al., 2018) or even a hypothesis to test for the project (e.g., stimuli database creation). Newer power and sample size planning suggestions include Accuracy in Parameter Estimation Maxwell, Kelley, \& Rausch (2008) and simulation of proposed analyses (Chalmers \& Adkins, 2020). These toolkits provide flexibility in traditional power analyses that focus on the if-this-then-that approach, yet, both AIPE and simulation require either a specific parameter (e.g., mean, effect size, etc.) or statistical test for planning sample size. In this tutorial, we explore how these latter two approaches can be combined to accommodate studies that may not have a specific hypothesis test or wish to account for the potential of a multiverse of analyses. Specifically, the examples focus on studies that implement multiple items and suggest that sample sizes can be planned to measure those items adequately and accurately, regardless of statistical test. Results show that pilot data can be used to determine a sample size that represents well-measured data, and multiple code vignettes will be provided for researchers to adapt and apply to their own measures.
}

@@

Statistical power and power analyses are arguably one of the most important components to planning a research study (Cohen, 1990). Yet, if reviews of transparency and openness in research publications are any clue, the social sciences have failed to fully implement power analyses as part of their common efforts Hardwicke et al. (2022). The replication ``crisis'' and credibility revolution have shown that many published studies in psychology are underpowered Vazire (2018). Pre-registration of a study involves outlining the study and hypotheses before data collection begins Nosek \& Lakens (2014) and then should summarily include a power analysis to determine the sample size necessary to detect the expected effect. Given the combined issues of publish-or-perish and that most non-significant results do not turn in published manuscripts, one may expect that power analysis would be especially critical for early career researchers Simmons, Nelson, \& Simonsohn (2011). Potentially, it is uninformative to publish studies that are underpowered (Halpern, 2002), but it can be difficult to know if a bad power analysis is better than no power analysis. A recent review of power analyses found in psychology journal articles indicates that researchers did not provide enough information to understand their power analyses and often chose effect sizes that were inappropriately justified (Beribisky, 2019).

One potential solution to the power analysis problem is the plethora of tools made available for researchers to simply power. G*Power is one of the most popular free power software options Erdfelder, Faul, \& Buchner (1996) that provides a simple point and click graphical user interface for power. Web-based tools have also sprung up for overall and statistical test specific sample size planning including powerandsamplesize.com and \url{https://designingexperiments.com} (Anderson, Kelley, \& Maxwell, 2017). Coding based packages, such as \emph{pwr} (Champely et al., 2017), \emph{faux} (DeBruine, 2021), and \emph{SimDesign} (Chalmers \& Adkins, 2020) can be used to examine power and sample size planning using \emph{R}, usually with simulation. Researchers have to be careful using any toolkit, as errors can occur with the over-reliance on software (Nuijten, Hartgerink, Assen, Epskamp, \& Wicherts, 2016). When computing sample size estimates it is important to remember that these values are estimations, not exact calculations guaranteed to produce a specific result Batterham \& Atkinson (2005).

Changes in publication practices and research design have also shown a new wrinkle to providing a sample size plan for a research study. While statistics courses often suggest that a specific research design leads to a specific statistical test, multiple Many Analysts papers have shown that - given the same data and hypothesis - researchers can come up with multiple ways to analyze the data Coretta, Casillas, {[}participating authors{]}, \& Roettger (n.d.). Research projects often have multiple testable hypothesis, however, it is unclear which hypothesis or test the sample size planning should be performed on. Further, an entire set of common research publications may not even have a hypothesis to examine within their project, as they are simply providing a large, quality dataset for future reuse {[}i.e., stimuli database creation (Buchanan, Valentine, \& Maxwell, 2019). The increased ability to compute complex statistical analyses, such as multilevel modeling, has pushed researchers with repeated measures designs to abandon creating person-level averages just to be able to use a traditional ANOVA (Brysbaert \& Stevens, 2018). These analyses have also made it clear that we should be careful to assume that all items in a research study have the same ``effect'', as there is often variability in their impact on the outcomes of the study.

In this manuscript, we show a proposed method to account for variability in item effects, a potential lack of hypothesis test (or simply not a good way to estimate an effect size of interest), and/or an exploratory design with an unknown set of potential hypotheses and analyses choices. These methods are inspired by newer sample size planning methods including Accuracy in Parameter Estimation {[}AIPE, @kelley2007, @maxwell2008{]} and the ability to simulate proposed data for item estimates (Rousselet, Pernet, \& Wilcox, n.d.). AIPE shifts the focus away from finding a \emph{significant} \emph{p}-value to finding a parameter that is ``accurately measured''. For example, a researcher may wish to detect a specific sized correlation in a study, \emph{r} = .35. They could then use AIPE to estimate the sample size needed to find a ``sufficiently narrow'' confidence interval around that correlation. Sufficiently narrow is often defined by the researcher using a minimum parameter size of interest and confidence intervals. Therefore, they could decide that their 95\% confidence interval should be approximately {[}.20, .50{]}, and sufficiently narrow was defined as a width of .30 or .15 on each side. While confidence intervals are related to Null Hypothesis Significance Testing (i.e., 95\% confidence intervals that do not include zero would indicate a significant difference from zero at \(\alpha\) \textless{} .05), AIPE procedures instead suggest a sample size that should obtain that width of a confidence interval, regardless if it includes zero.

In this approach, a researcher could use sequential testing to estimate their parameter of interest after each participant to determine if they have achieved their expected width of the confidence interval around that parameter. One would set a minimum sample size (i.e., based on known data collection ability), use the confidence interval width as a stopping rule (i.e., stop data collect when the CI is narrow), and use the estimated sample size from the AIPE calculations as a potential maximum sample size. By defining each of these components, a research could ensure a feasible minimum sample size, a way to cease data collection when goals have been met, and a stopping rule to ensure an actual end to data collection. Given pilot or previously collected data, we should be able to leverage the ideas behind AIPE, paired with simulation and bootstrapping, to estimate the minimum and maximum proposed sample sizes and stopping rules for repeated measures studies with expected variabillity in parameter estimates for items.

@@

Using these ideas, we suggest the following procedure to determine a sample size for each item:

@@

\emph{Samples}. Each population was then sampled as if a researcher was conducting a pilot study. The sample sizes started at 20 participants per item increasing in units of 10 up to 100 participants.

@@

In this section, we simulate what a researcher might do if they follow our suggested application of AIPE to sample size planning based on well measured items. Assuming each pilot sample represents a dataset a researcher has collected, we will simulate samples of 20 to 2000 increasing in units of 20 to determine what the new sample size suggestion would be. We assume that samples over 500 may be considered too large for many researchers who do not work in teams or have participant funds; however, the sample size simulations were estimated over this amount to determine the pattern of suggested sample sizes (i.e., the function between original sample size and proposed sample size). The standard error of each item was calculated for each suggested sample size by pilot sample size by population type.

@@

In order to minimize any potentially random quirks, we simulated the sample selection from the population 100 times and the researcher simulation 100 times for each of those selections, resulting in 10000 simulations of all combinations of variables (i.e., scale of the data, heterogeneity, pilot study size, researcher simulation size). The average of these simulations is presented in the results.

@@

\(N_{Pilot}\) indicates the sample size of the pilot data minus the minimum simulated sample size to ensure that the smallest sample sizes do not decay (i.e., the formula zeroes out). This value is raised to the power of \(log_2\) of the sample size of the pilot data, which decreases the impact of the decay to smaller increments for increasing sample sizes. This value is then multiplied by the projected sample size. As shown in Figure \ref{fig:corrected-figure}, this correction factor produces the desired quality of maintaining that small pilot studies should \emph{increase} sample size, and that sample size suggestions level off as pilot study data sample size increases.

@@

To explore if we could recover the new projected sample size from data a researcher would have, we used linear models to create a formula for researcher correction. First, the corrected projected sample size was predicted by the original projected sample size. Next, the standard deviation of the item standard deviations was added to the equation to recreate heterogeneity estimates. The scale of the data is embedded into the standard deviation of the items (\emph{r} = 0.80), and therefore, this variable was not included separately. Last, we included the pilot sample size.

The first model using pilot sample size to predict new sample size was significant, \emph{F}(1, 2266) = 23,280.26, \emph{p} \textless{} .001, \(R^2 = .91\), 90\% CI \([0.91, 0.92]\), capturing nearly 90\% of the variance, \(b = 0.53\), 95\% CI \([0.52, 0.54]\). The second model with item standard deviation was better than the first model \emph{F}(1, NULL) = 13.59, \emph{p} \textless{} .001, \(R^2 = .91\), 90\% CI \([0.91, 0.92]\). The item standard deviation predictor was significant, \(b = 0.01\), 95\% CI \([0.00, 0.02]\), \(t(2265) = 2.20\), \(p = .028\). The addition of the original pilot sample size was also significant, \emph{F}(1, NULL) = 4,101.10, \emph{p} \textless{} .001, \(R^2 = .97\), 90\% CI \([0.97, 0.97]\).

@@

\begin{tabular}{lllll}
\toprule
Term & Estimate & $SD$ & $t$ & $p$\\
\midrule
Intercept & 34.549 & 0.425 & 81.264 & < .001\\
Projected Sample Size & 0.621 & 0.003 & 247.039 & < .001\\
Item SD & 0.000 & 0.003 & 0.014 & .989\\
Pilot Sample Size & -0.483 & 0.008 & -64.040 & < .001\\
\bottomrule
\end{tabular}

@@

Last, we examine the question of an appropriate SE decile. All graphs for power, heterogeneity, scale, and correction are presented online. First, the 0\%, 10\%, and 20\% deciles are likely too restrictive, providing very large estimates that do not always find a reasonable sample size in proportion to the pilot sample size, scale, and heterogeneity. If we examine the \(R^2\) values for each decile of our regression equation separately, we find that the 50\% (0.97) represents the best match to our corrected sample size suggestions. The 50\% decile, in the corrected format, appears to meet all goals: 1) increases with heterogeneity and scale of data, and 2) higher suggested values for small original samples and a leveling effect at larger pilot data. Figure \ref{fig:decile-figure} illustrates the corrected scores for simulations at the 50\% decile recommended cutoff for item standard errors.

@@

In this section, we provide two examples of the suggested procedure. The first example includes concreteness ratings from Brysbaert, Warriner, and Kuperman (2014). Instructions given to participants denoted the difference between concrete (i.e., ``refers to something that exists in reality'') and abstract (i.e., ``something you cannot experience directly through your senses or actions'') terms. Participants were then asked to rate concreteness of terms using a 1 (abstract) to 5 (concrete) scale. This data represents a small scale dataset that could be used as pilot data for a study using concrete word ratings. The data is available at \url{https://osf.io/qpmf4/}. The second dataset includes a large scale dataset with response latencies, the English Lexicon Project (Balota et al., 2007). The English Lexicon Project consists of lexical decision response latencies for English words. In a lexical decision task, participants simply select ``word'' for real words (e.g., \emph{dog}) and ``nonword'' for pseudowords (e.g., \emph{wug}). The trial level data is available here {[}\url{https://elexicon.wustl.edu/}{]}. Critically, in each of these examples, the individual trial level data for each item is available to simulate and calculate standard errors on. Data that has been summarized could potentially be used, as long as the original standard deviations for each item were present. From the mean and standard deviation for each item, a simulated pilot dataset could be generated for estimating new sample sizes. All code to estimate sample sizes is provided on our OSF page.

@@

The concreteness ratings data includes 63039 concepts that were rated for their concreteness. In our fictional study for this example, we selected 100 random words to show participants. In the original study, not every participant rated every word, which created uneven sample sizes for each word. In our random sample of 100 words, the average pilot sample size was 28.15 (\emph{SD} = 1.59), and we will use 28 as our pilot sample size for this example. All ``do not know'' ratings were set as missing data. The 50\% decile for items standard error was 0.25 for our cutoff criterion.

The pilot data was then simulated, with replacement, with samples from 20 to 300 increasing in units of 5. On each sample, the percent of items below the cutoff score were calculated. After applying our correction equation, we find that a sample size of 44 would allow for at least 80\% of items to meet the cutoff criterion. The sample sizes for 85\% (48), 90\% (48), and 95\% (51) are also options for sample size suggestions. Finally, we calculated the potential amount of data retention given that participants could indicate they did not know a word (\(M_{correct}\) = 0.79, \emph{SD} = 0.24). In order to account for this facet, the potential sample sizes were multiplied by 1/0.79, which results in a suggested sample of 56, 61, and 65. Therefore, we could designate our minimum sample per item as 56, stopping rule of 0.25, and maximum sample size of 65.

@@

The ELP response latency data includes 80962 word-forms, 40481 that are listed as non-words, and 40481 real words. For our example study, we will randomly select 500 real words and 500 non-words to show participants. The average pilot sample size for this random sample was 32.72 (\emph{SD} = 0.64), and \emph{n} = 33 will be our pilot size for this example. Again, participants are expected to make mistakes, and we calculated percent correct as 0.85, which was roughly even in the two stimulus categories: \(M_{word}\) = 0.82 and \(M_{non-word}\) = 0.88. The 50\% decile for items standard error was 61.23 for our cutoff criterion. We additionally checked to ensure that the two stimulus types did not have very different cutoff criterions: 50\% decile \(SE_{words}\) = 58.98, 50\% decile \(SE_{nonwords}\) = 62.61. In this scenario, we could choose to go with the lower SE to be more conservative (i.e., higher projected sample size). Given the values were close for large scale data, we used the 50\% decile of all stimuli taken together.

The pilot response latency data was then simulated in the same way as described above. After calculating the percent below our cutoff score, we applied the correction to the projected sample sizes. A sample size of 31 would equate to 80\% of the items reaching our cutoff, along with 85\% (34), 90\% (34), and 95\% (38). Again, we adjusted for data loss given that participants are expected to incorrectly answer items, resulting in a suggested sample of 36, 40, and 45. One other possible consideration for this study is potential fatigue in showing participants 1000 target items. Therefore, we could designate in our research design that each participant will only receive 500 of the target items. We would need to double our sample sizes to account for splitting of the items across multiple sets of participants. Our minimum sample size for the entire study could be 72, stopping rule of 61.23, and maximum sample size of 90. This study would benefit from an adaptive design, where smaller sets items are randomly sampled for participants until they reach the minimum sample size or the cutoff criteria. At this point, items are probabilistically sampled (e.g., higher selection probability for items that have not reached a minimum or stopping rule) until all items have reached criteria.

@@

While the examples in this manuscript are traditionally cognitive linguistics focused, any research using repeated items can benefit from newer sampling techniques. Therefore, we provide XX example vignettes and code examples on our OSF page/GitHub site for this manuscript across a range of examples of data types provided by the authors of this manuscript. Examples include psycholinguistics, social psychology, COVID related data, and cognitive psychology.

@@

In this manuscript, we demonstrated a method using AIPE and simulation/bootstrapping to estimate a minimum and maximum sample size along with a rule for stopping data collection based on narrow confidence intervals on the parameter of interest. We believe this procedure is specifically useful for studies with multiple items that intend on using item level focused analyses; however, the utility of measuring each item well can extend to many analysis choices. By focusing on gathering quality data, we can suggest that the data is useful, regardless of outcome of any hypothesis test.

One limitation of these methods would be using datasets with very large numbers of items to simulate what might happen within one study. For example, the English Lexicon Project includes thousands of items, and by the time we would simulate for all of those, it would likely suggest needing thousands of participants for \emph{most items} to reach criterion. Alternatively, as the number of items increases, you also could potentially see very small estimates for sample size due to the correction factor (as with large numbers of items, you could find many items with standard errors below the 50\% decile). Therefore, it would be beneficial to consider only simulating with what a participant would reasonably complete in a study. On the other side, small numbers of repeated items usually result in higher sample sizes proposed from the original pilot data. This result occurs because the smaller number of items means more samples for nearly all to reach the cutoff criteria. These results are not too different than what we might expect for a power analysis using a multilevel model - larger number of items tends to decrease necessary sample size, while smaller numbers of items tend to increase sample size.

Second, these methods do not ensure the normal interpretation of power, wherein you know would find a specific effect for a specific test, \(\alpha\), and so on. As discussed in the introduction, there is not necessarily a one-to-one mapping of hypothesis to analysis, and many of the guesses within a traditional power analysis are just that - best guesses for various parameters. These methods could be used together to strengthen our understanding of sample size necessary for both a hypothesis test and well tuned estimation.
